---
title: The spells die rolls plot
output:
  litedown::html_format:
    meta:
      css: ["@article", "../_liteblog.css"]
      js: ["@sidenotes"]
      include-before: "_liteblog-header.html"
      include-after: "_liteblog-footer.html"
    options:
      toc: true
      js_highlight:
        package: prism
        style: tomorrow
---

```{r}
#| label: setup
#| echo: false
.blog <- readRDS("../_liteblog.rds")
```

On this page I'll walk through the code used to produce the "spell dice" plot that I posted on social media. It's not the most exciting code I've ever written but it was useful as a tiny side project that I could use to teach myself targets. Note that this post doesn't use targets to render the plot, except in the trivial sense the entire blog is built with targets. Within the post, it's just regular R code. Speaking of which, I suppose I'd best load the packages that the plot relies on:

```{r}
#| label: load-packages
#| message: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)
library(readr)
library(forcats)
library(ggrepel)
```

## Read data

The `spells` data that I'm using here comes from the [TidyTuesday D&D Spells](https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-12-17) data set. The data set was compiled by Jon Harmon, and originates in the recently released Dungeons & Dragons Free Rules (2024 edition). If you've played D&D before, this should be quite familiar:

```{r}
#| label: read-spells
#| print: print_tinker
spells <- read_csv("./data/spells.csv", show_col_types = FALSE)
print(spells)
```

## Tidy data

```{r}
#| label: tidy-data
#| print: print_tinker
dice_dat <- spells |>
  select(name, level, description) |>
  mutate(
    dice_txt = str_extract_all(description, "\\b\\d+d\\d+\\b"),
    dice_txt = map(dice_txt, unique)
  ) |>
  unnest_longer(
    col = "dice_txt",
    values_to = "dice_txt",
    indices_to = "position"
  ) |>
  mutate(
    dice_num = dice_txt |> str_extract("\\d+(?=d)") |> as.numeric(),
    dice_die = dice_txt |> str_extract("(?<=d)\\d+") |> as.numeric(),
    dice_val = dice_num * (dice_die + 1)/2,
    dice_txt = factor(dice_txt) |> fct_reorder(dice_val)
  )

print(dice_dat)
```

## Make plot

```{r}
#| label: make-plot
#| fig.width: 12
#| fig.height: 6
palette <- hcl.colors(n = 10, palette = "PuOr")

labs <- dice_dat |>
  summarise(
    dice_txt = first(dice_txt),
    count = n(),
    .by = dice_txt
  )

pic <- ggplot(
  data = dice_dat,
  mapping = aes(
    x = dice_txt,
    fill = factor(level)
  )
) +
  geom_bar(color = "#222") +
  geom_label_repel(
    data = labs,
    mapping = aes(
      x = dice_txt,
      y = count,
      label = dice_txt
    ),
    size = 3,
    direction = "y",
    seed = 1,
    nudge_y = 4,
    color = "#ccc",
    fill = "#222",
    arrow = NULL,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    name = "Spell level",
    values = palette
  ) +
  scale_x_discrete(
    name = "Increasing average outcome \u27a1",
    breaks = NULL,
    expand = expansion(.05)
  ) +
  scale_y_continuous(name = NULL) +
  labs(title = "Dice rolls in D&D spell descriptions by spell level") +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "#222"),
    text = element_text(color = "#ccc"),
    axis.text = element_text(color = "#ccc"),
    axis.title = element_text(color = "#ccc"),
    plot.margin = unit(c(1, 1, 1, 1), units = "cm"),
    legend.position = "inside",
    legend.position.inside = c(.3, .825),
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.byrow = TRUE
  )

plot(pic)
```

