---
title: "A deep dive into table1"
description: "This is a subtitle"
date: "2024-06-21"
--- 

<!--------------- my typical setup ----------------->

```{r}
#| label: setup
#| include: false
very_wide <- 500
wide <- 136
narrow <- 76
options(width = narrow)
cache_images <- TRUE
set.seed(1)
```

<!--------------- post begins here ----------------->


```{r}
#| label: packages
#| message: false
#| warning: false
library(palmerpenguins)
library(table1)
library(dplyr)
```

The [table1](https://benjaminrich.github.io/table1/) package ([github link](https://github.com/benjaminrich/table1))

Basic usage:

```{r}
#| label: minimal-use
table1(~ island + bill_length_mm | species, penguins)
```

## Labels

Adding labels:

```{r}
#| label: labelling-1
label(penguins$island) <- "Island"
label(penguins$bill_length_mm) <- "Bill Length (mm)"

table1(~ island + bill_length_mm | species, penguins)
```

Using the `label<-()` function in table1 is how I've seen most people do it, but in truth there's nothing fancy here: all that the function does is set the "label" metadata attribute for the relevant object, so you could totally do it in base R via the `attr<-()` function:

```{r}
#| label: labelling-2
attr(penguins$bill_depth_mm, "label") <- "Bill Depth (mm)"

table1(~ island + bill_depth_mm | species, penguins)
```

However, I am slowly coming to like the `setLabel()` convenience function in table1 because it returns the labelled object and hence plays nicely with a dplyr workflow: 

```{r}
#| label: labelling-3
penguins <- penguins |> 
  mutate(
    flipper_length_mm = setLabel(flipper_length_mm, "Flipper Length (mm)"),
    body_mass_g = setLabel(body_mass_g, "Body Mass (g)"),
    sex = setLabel(sex, "Sex"),
    year = setLabel(year, "Year")
  )

table1(~ flipper_length_mm + body_mass_g + sex + year | species, penguins)
```

## Custom statistics

Okay so yeah you can do this to get geometric means and standard deviations...

```{r}
#| label: custom-render-1
table1(
  x = ~ flipper_length_mm + body_mass_g + sex + year | species, 
  data = penguins, 
  render.continuous = c("Geometric mean" = "GMEAN", "Geometric SD" = "GSD")
)
```

That's important in my job where almost every continuous variable is lognormal. You can see that this affects only the continuous variables not the discrete ones. Similarly you can do this...

```{r}
#| label: custom-render-2
table1(
  x = ~ flipper_length_mm + body_mass_g | species, 
  data = penguins, 
  render.continuous = c(
    "10th percentile" = "q10", 
    "50th percentile" = "q50",
    "90th percentile" = "q90"
  )
)
```

The tricky part to this is that it's not documented. You have to dig into the [source code](https://github.com/benjaminrich/table1/blob/master/R/table1.R) to work out what "aliases" are supported.
